---
#layout: preprint
layout: review, 11pt
title: "RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig(at)gmail.com
    footnote: Corresponding author
  - name: Hilmar Lapp
    affiliation: NESCent
  - name: Scott Chamberlain
    affiliation: ropensci
  - name: Rutger Vos
    affiliation: NBC
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center, Durham, NC, USA
  - code: ropensci
    address: |
      University of California, Berkeley, CA, USA
  - code: NBC
    address: | 
      Naturalis Biodiversity Center, Leiden, the Netherlands
abstract: | 
      1. NeXML is a new phylogenetic informatics data and metadata exchange 
      format that offers substantial improvements in reliability, extensibility, 
      and richness over earlier standards such as NEXUS. In this paper we present 
      the RNeXML package which provides a bridge between this rich and reliable
      data format and the extensive library of phylogenetic tools already
      available in R, with a particular emphasis on comparative methods.

      2. RNeXML is much richer than a basic parser: here we illustrate
      how the package facilitates adding and manipulating semantic 
      annotations, that is annotations whose meaning can be defined externally
      by way of semantic web technologies.

      3. An essential feature of NeXML is to provide both the flexibility
      to be continually extended to meet the emerging needs of the phylogenetics
      community while remaining backwards-compatible with software developed
      for earlier versions.  Here we illustrate how RNeXML makes it easy
      to extend NeXML in this manner.  
      
      4. RNeXML is well suited for users to preserve and maintain their
      personal archives of phylogenetic and comparative trait data in consistent,
      query-able, and cross-platform format, to leverage the rapidly growing 
      public community archives of rich NeXML files, and to interact with web
      services that deliver their results (with their provenance) as NeXML.

bibliography: components/references.bib
csl: components/methods-in-ecology-and-evolution.csl
documentclass: components/elsarticle

output: 
  pdf_document:
    template: components/elsarticle.latex
    keep_tex: true
    fig_caption: true


---

<!------------------------------------------------

Map/Outline:
------------

- Introduction
  - Exciting times in phylogenetics!  
  - Challenges with existing formats 
  - Overview of NeXML
  - Map 

- Tutorial
  - Reading Phylogenetic trees
  - Reading Character data
  - Writing phylogenetic trees & character data

  - Metadata
    - Adding basic metadata
    - Adding taxonomic identifiers
    - Adding Generic metadata
    - Parsing metadata. XML. RDF.
    - Extending NeXML with metadata: Example defining simmap stochastic character mappings.  

  - Publishing NeXML 

- Implementation, Quality Control
- Software reuse and support

------------------------------------------------>





```{r compile-settings, include=FALSE}
library("methods")
library("knitr")
opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
               cache = 1, comment = NA, verbose = TRUE)
basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile')) 
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))


```

# Introduction



Though R provides the most extensive and widely used collection of
comparative phylogenetic methods tools, R users lack a convenient,
extensible format to exchange and archive comparative methods data.
A common file format for sharing phylogenetic tree data has long played
an essential role in the advancement of phylogenetics research.
The rapid increase in phylogenetics data and research methods that make
use of such data has introduced many new challenges and opportunities 
to sharing this information both between software environments and 
between research groups [@Drew_2013; @Stoltzfus_2012].  

<!-- FIXME 

ADD STUFF ABOUT:

- Rapid growth in phylogenetics data (+citations)
- Rapid growth of research methods and software that use phylogenetic data. Comparative methods in evolution. Ecological phylogenetics. Biodiversity informatics. (+citations)
- More on data sharing challenges and opportunities a la @Drew_2013
- More on the hoops that users need to jump through if trees, data
and other things all need to come from separate files and need to be
joined on taxon labels.

-->

The ever expanding needs of the phylogenetics community are rapidly
out-pacing the capacity of current standards, which is resulting in a
radiation of different data representations and a loss of interoperability
at the very time when the growth of available data and methods has made
that interoperability most valuable. 

For example, the currently popular NEXUS file format cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair).  This has led developers to
repeatedly add small modifications to the standard to meet the needs
of their own software, which results in mutually incompatible dialects
that largely defeat the advantages of a shared  file format.
NeXML was developed to addresses each of these problems [@Vos_2012]  
by drawing on robust web standards in data representation.
The design and expressiveness of NeXML makes it much easier to meet
the rising bar for best practices in data sharing in phylogenetic 
research [@Cranston_2014].  



<!-- What do we need to say, if anything, about the current R formats?  about phylobase?  -->

The R language [@R] has become an important environment for 
phylogenetics research, with over 30 packages available including several
with over 1000 citations [@taskview],
though stand-alone programs remain an important part of this ecosystem
for tasks such as generating phylogenetic trees from sequence data.  Within 
this ecosystem a variety of different data formats for
phylogenetic data have been developed and modified. Most but not all 
are based on the simple and popular `ape::phylo` format provided by the
ape package [@Paradis_2004]. 

Though certain functions allow R users to switch between these formats
within R, these formats frequently contain more information than their
NEXUS file counterparts, and users lack a suitable alternative for
archiving and sharing phylogenetic data outside of the R ecosystem. 
Consequently, information added to phylogenies by these R programs cannot 
easily be shared, archived, and made available to other software;
but inherits the limitations of the NEXUS format.  

`RNeXML` opens up a rich world of both currently available and rapidly
growing data sets in the NeXML format to the R user and the large ecosystem
of R packages currently available for handling phylogenetic data  
R packages and R users frequently add rich information to the phylogenies
they work with, but need a platform independent data exchange format to
share and archive this. We illustrate how NeXML can
be used to distribute both phylogenies and character trait data used in 
many popular phylogenetic comparative methods tools in R.  
`RNeXML` also provides convenient tools for adding metadata
to NeXML files, including the automated generation of metadata and examples
of how to extend the NeXML standard to represent new data features without
breaking validation or backwards compatibility of the file type.  

<!-- FIXME something about how we will increasingly want to work with (semantic) metadata
and how RNeXML provides a solid platform from which to do so.  
-->

<!-- Map paragraph -->

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end with an example that uses metadata to
extend the NeXML standard to describe stochastic character mapping, one
of the kinds of data annotation that is currently available in R formats
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data repository directly from R.
Finally, we provide a brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

This paper has been written as a dynamic document 
[@Xiei_2013].  The reader may paste the code blocks
shown into their R terminal to follow along with the exmples shown.


# Introduction to RNeXML


A NeXML document can be thought of as having multiple components: (1) phylogeny
data (2) character-trait data, (3) operational taxonomic units, and (4) metadata. 
`RNeXML` seeks to provide a seamless way to convert these to and from their 
corresponding R object representations, such as `ape::phylo` phylogenies, character
matrices (`data.frames`) and metadata lists (or for advanced users,
`XMLInternalDocument` and `rrdf` triplestores). Metadata can provide
information about the study in which the phylogeny was generated or
applied (such as authors, citations to publications, timestamps, and even
descriptions of methods), but can also be used to annotate individual
data components.  We illustrate how metadata annotations can be added to
the operational taxonomic units of the phylogeny and can be used to extend
the data representation to new uses such as stochastic character maps.


## Getting Started


The current stable version of the `RNeXML` package can be installed from
the CRAN library[^1].  

```r
install.packages("RNeXML", dependencies=TRUE)
```


## Reading phylogenetic trees



```{r echo=FALSE}
library(RNeXML)
```

<!-- I still question the wisdom of having multiple names for the same
method. At first this might seem user-friendly, but it will just be confusing
and create a document and source code maintenance nightmare. Why not just pick
read.nexml and be done with it? -->
<!-- For whatever reason, R has native support for aliasing function names;
     so I don't see this creating a maintenance/documentation issue.  If I 
     had to pick, I'd go with read_nexml though; a lot of the R developer 
     community strongly opposes the `.` in function names, since it falsely
     mimics the S3 class structure.  - CB -->

We begin by loading the `RNeXML` package and reading in an example
NeXML file provided with the package, containing two small phylogenetic
trees.[^2]

```{r}
library("RNeXML")
f <- system.file("examples", "trees.xml", package="RNeXML")
nex <- nexml_read(f)
```

The resulting `nex` object is an R object representation of the NeXML. This is 
a native R object, of the `nexml` class, for representing phylogenetic trees (as well as 
character traits, and metadata), analogous to the `ape::phylo` object from the 
popular `ape` R package [@Paradis_2004]. While R functions and packages 
could be written to use this `nexml` object directly, most users will prefer to use
already existing methods provided by packages they know. From a `nexml` object, we can 
therefore extract any phylogenies it contains in the `ape::phylo` format. In this 
case we will get back a `multiPhylo` object, since the file contains two phylogenies.  

```{r}
phy <- get_trees(nex)
```

We can then leverage the rich suite of methods available from the R
phylogenetics community. For instance, here we use the standard plotting function from
the `ape` package to plot the first tree in the resulting `ape::phylo` object returned by
`get_trees`.

```{r eval=FALSE}
plot(phy[[1]])
```

NeXML can not only contain multiple phylogenetic trees in a single document,
but can also group sets of trees into separate blocks (e.g. as resulting from
different workflows). The `get_trees` function attempts to return the most sensible 
R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
object; if multiple trees are available in a single `trees` block, it will return
an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
blocks are available, it will return a list of `multiPhylo` blocks.
While most users will find this behavior convenient, there are times
(such as looping over many NeXML files) when having a variable return
type is undesirable.  To always get back a list of `multiPhylo` objects,
even when only one tree is available, use `get_trees_list()`.

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user.  For instance,
a user can access all the phylogenies available in TreeBASE <!-- first mention -->
<!-- of TreeBase, include URL? or introduce it in intro? --> through the
`nexml` format.  `RNeXML` can read directly from a URL:

```{r eval=FALSE}
nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
```

Previously, this was possible using the `treebase` package [@Boettiger_2012"]
for R, which relied on the NEXUS parser provided in the `ape` package.
Due to inconsistencies present in the NEXUS standard [@Vos_2012], 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Using the more precisely defined
NeXML standard, every valid NeXML document on TreeBASE can now successfully
be parsed to R objects.



## Reading character data 

Like NEXUS files, NeXML documents can also contain character data.  While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence (or morphological) data 
used to infer the phylogenetic tree, we find this ideal for character data used in
comparative methods popular in R. Methods to extract character data
work much like those for the phylogenetic data. Here we read in an example
NeXML file that contains both a phylogeny and character data. 
`get_characters` returns a `data.frame` with columns as characters and rows as
taxa:

```{r }
comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
nex <- nexml_read(comp_analysis)
get_characters(nex)
```

As this example shows, continuous and discrete characters are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. The `get_characters_list()` function will consistently receive
a list of data frames corresponding to each character matrix block of
the NeXML.


## Writing phylogenetic trees 

Generating NeXML files from R is likewise straightforward.  Here we
write a phylogeny from the `ape::phylo` format out to a NeXML file:

```{r}
data(bird.orders)
nexml_write(bird.orders, file = "birds.xml")
```

<!-- in caper there is this method comparative.data to join data
and trees on taxon labels. it would be nice if this merged data
structure could also be passed to nexml_write (or should that be
write.nexml?) -->
<!-- Great observation.  I think we should add a coercion method
from the caper structure to the nexml S4 structure.  Then nexml_write
can automatically handle this. added as issue now  -->

The first argument to `nexml_write` can be a `ape::phylo` phylogeny,
`multiPhylo` list of phylogenies, or a `nexml` object -- anything that
can be coerced into the `nexml` class.  As we have seen, NeXML files
can contain multiple phylogenetic trees, as well as character matrices.
We can add both trees  and character data using `nexml_write`,

```{r}
library("geiger")
data(geospiza)
nexml_write(trees = geospiza$phy, characters = geospiza$dat, file="geospiza.xml")
```

These examples use data from the popular phylogenetics software `geiger` [@Harmon_2008].  

<!-- how about add_taxa? -->
<!-- A taxa block is added when trees or characters are added.  Yeah,
     I guess it would make sense to be able to add taxa directly,
     (e.g. a NeXML file containing only RDFa metadata about taxa, 
     independent of characters or trees), but I worry about confusing
     people into thinking they must add taxa blocks separately, and
     about when we should merge or not merge taxa blocks of matching
     species. See https://github.com/ropensci/RNeXML/issues/43 
     Added new issue for this, https://github.com/ropensci/RNeXML/issues/85 -->

NeXML files can also be built up bit by bit using the `add_characters`,
`add_trees`, and `add_meta`.  Each of these functions can add data to an
existing `nexml` object or create a new nexml object if none is provided.
Here we first create a `nexml` object containing the phylogeny data,
and then add the character data to it:

```{r}
geiger_nex <- add_trees(geospiza$phy)
geiger_nex <- add_characters(geospiza$dat, geiger_nex)
```

The data need not share the same taxa.  We can append additional
phylogenies and characters corresponding to different taxa:

```{r}
data(primates)
geiger_nex <- add_trees(primates$phy, geiger_nex)
geiger_nex <- add_characters(primates$dat, geiger_nex)
```

This allows us to include multiple trees and character matrices to a single NeXML file.  

# Metadata

At the heart of the NeXML standard is the ability to encode arbitrary
metadata in precisely specified, machine-readable formats that maintain
backwards compatibility with any existing NeXML parser. NeXML uses the
"Resource Description Framework in Annotations", or RDFa,
as described in @Vos_2012 or @W3C_2014. 
These annotations can be used to add 
information to any particular element of the data, (a tree, node, edge,
or taxonomic unit; a character matrix or specific trait, or even another 
metadata annotation or the NeXML data file as a whole). `RNeXML` provides
a range of tools for accessing metadata embedded in a NeXML file.  



## Writing NeXML metadata

Researchers cannot take advantage of NeXML metadata manipulation if
there is no metadata there.  The burden of adding good metadata remains a
significant barrier to increased sharing and reuse of scientific research
[@Tenopir_2011].
`RNeXML` seeks to reduce the burden of adding
high quality, machine-readable metadata to NeXML files. This occurs on
several levels:

- Automatic annotation
- Adding common metadata fields (`add_basic_metadata`)
- Taxonomic identifiers (`taxize_nexml`)
- Generic metadata (`add_meta`)
- Extending the NeXML standard using RDFa

<!-- in the Perl library I also embed the value of the environment variable
$USER as a document-level annotation -->
<!-- Good idea-->

`RNeXML` automatically adds basic metadata by default such as a time-stamp
of when the tree was produced and an open data license (CC0, compatible
with deposition on Dryad data repository), which can be configured
as needed.  

To adjust these and other common metadata fields that should 
be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
function.  This function does not assume any familiarity with the concepts
of namespaces or other more advanced metadata tools discussed later.  The
`add_basic_meta` takes an existing `nexml` object (or will create a new one
if none is provided, like the other `add_` functions), and any of the following
fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
`citation`.  Additional fields may be added to subsequent versions, see
`?add_basic_metadata` for the most up-to-date information.  Here we create
a new `nexml` object with the basic metadata for the `bird.orders` data
provided in the `ape` package:

```{r}
data(bird.orders)
birds <- add_trees(bird.orders)
birds <- add_basic_meta(birds,
  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",

  description = "This data set describes the phylogenetic relationships of the
     orders of birds as reported by Sibley and Ahlquist (1990). Sibley
     and Ahlquist inferred this phylogeny from an extensive number of
     DNA/DNA hybridization experiments. The ``tapestry'' reported by
     these two authors (more than 1000 species out of the ca. 9000
     extant bird species) generated a lot of debates.

     The present tree is based on the relationships among orders. The
     branch lengths were calculated from the values of Delta T50H as
     found in Sibley and Ahlquist (1990, fig. 353).",

  citation = "Sibley, C. G. and Ahlquist, J. E. (1990) Phylogeny and
     classification of birds: a study in molecular evolution. New
     Haven: Yale University Press.",

  creator = "Sibley, C. G. and Ahlquist, J. E.")
```

Citations can also be added using R's `bibentry` type, which allows
a user to add citations to R packages, 

```{r}
birds <- add_basic_meta(birds, citation = citation("ape"))
```

or to papers by simply using the object's DOI (using @Boettiger_2014).
For instance to add the citation information of the paper that generated the
`geospiza` phylogeny included in the `geiger` package: 

```{r}
library("knitcitations")
geiger_nex <- add_basic_meta(geiger_nex, citation = bib_metadata("10.2307/2408428"))
```



## Taxonomic identifiers

`RNeXML` provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.
One common use case is that of taxonomic identifiers.  Misspellings or
inconsistent use of taxonomic names in phylogenetic data are a common
source of frustration in reusing phylogenetic trees, especially when
integrating data from different sources: trees might come from one place
that uses one taxonomy, character data and occurrence data might come from
other places that use different taxonomies. Hence, not just misspellings,
but also synonyms, homonyms and incompatible taxon concepts might plague
the end user. By associating the taxonomic labels against a specific,
centralized database we can catch some of these inconsistencies and
indicate more precisely which classification we are appealing to in the
event of conflicting taxonomic definitions.  In this example, we indicate
that we intend the bird orders listed in the tree to correspond to the
NCBI definitions.

The `taxize_nexml` function uses the R package `taxize` 
[@Chamberlain_2013] to check each taxon label against the NCBI database.
If a unique match is found, a metadata annotation is added to the taxon
providing the NCBI identification number to the taxonomic unit.  If no
match is found, the user is warned to check for possible typographic
errors in the taxonomic labels provided.


```{r message=FALSE, results='hide'}
birds <- taxize_nexml(birds, "NCBI")
```


<!-- what if multiple NCBI taxon IDs are found, e.g. 'Aotus'? -->


## Custom metadata extensions

So far our examples have relied on functions that have a built-in
understanding of the kind of metadata being provided, and can thus
generate the appropriate metadata with little further guidance.
In order to be machine-readable and understandable, NeXML requires
all metadata annotations follow the RDFa format (see @Vos_2012) which
explicitly declares the meaning of the annotation. 

Some basic definitions about the XML syntax used to express NeXML
will be helpful for this discussion, though a more comprehensive 
introduction can be found in @Vos_2012.  For instance, consider how
the title we gave to the `bird.orders` phylogeny appears in the NeXML file
using the XML format:

```xml
<meta id="m2" property="dc:title" datatype="xsd:string"
      content="Phylogeny of the Orders of Birds From Sibley and Ahlquist"/> 
```

The format begins with an XML "tag" and the name "meta", indicated by the opening
angle bracket; `<meta`.  A closing tag `\>` indicates the end of this
XML "element" .  This tag contains several "attributes", named "id",
"property", etc., that an R user can think of as names in a `list`. What
follows the equals sign after the attribute name is called the attribute "value",
(e.g. the `id` attribute has value "m2").  

The name of any element (like "meta") or attribute (like "id") in XML must be associated
with a particular dictionary called a "namespace" and specified using a "prefix"
of an abbreviation followed by a colon appearing before the name.  If no prefix
is given, as is the case for all attributes here: `id`, `property`, `datatype`,
and `content`, then a default prefix specified at the beginning of the document applies,
in this case, the NeXML namespace.  This allows users to combine different XML
that have different definitions of terms like `id` or `property`.  

The example element above also uses namespaces on some of the attribute values themselves.
This is a consequence of RDFa annotation, and not strictly part of XML, though
it works much the same way.  While we hope the following examples will make these
concepts clearer, users do not need to understand XML and namespaces to use most
of the basic annotation features provided by `RNeXML`.   

So much for the structure; we can now briefly explain the meaning of this element. 
We see the title appears as the value of the `content` attribute of this `meta` element.
The `id` attribute provides us with a unique way to refer to this meta
element in this document.  The `datatype` tells us that the content
is a text string (as opposed to, say, a numerical or boolean value).
The `property` is the attribute that most interests us here; it tells us
that the content is a title.  The prefix `dc` indicates the vocabulary
(here identified as an XML namespace) to which the word `title` belongs.  The prefix
is defined elsewhere in the NeXML file, as we can see with:


```{r}
prefixes <- get_namespaces(birds)
prefixes["dc"]
```

This URL links us to a web page providing both a human and machine-readable
version of precisely what the term `title` means, showing us that it
refers to `title` in the sense of a creative work, rather than the title
of a person, a legal title, the verb title, etc.  This precision can allow
us to perform precise computer reasoning on metadata, without relying on
heuristics or assumptions, as we illustrate later.  For the moment though,
the need to refer to an explicit controlled vocabulary (or ontology)
with a url, saying "dc:title" rather than just saying the property is
"title", is something of a burden in creating metadata.  The previous
examples involve functions that already know what kind of metadata is
being provided (a title, a data creator, etc), and so can provide the
appropriate vocabulary automatically.  The main advantage of linking to
an external vocabulary, however, is the ability to express concepts that
are not already "built in" to the data model.  `RNeXML` provides tools
to generate arbitrary metadata annotations given an external vocabulary,
such as we illustrate here.


From the list of namespaces in `prefixes`, we can follow links a handful
of established vocabularies and find terms that are not included in
`add_basic_meta`.  For instance, by resolving the location identified
by the `prism` prefix,

```{r}
prefixes["prism"]
```

we will find that there is a term for `modificationDate` for a file.  We create
a `meta` element containing this annotation using the `meta` function:

```{r}
modified <- meta(property = "prism:modificationDate", content = "2013-10-04")
```

We can add this annotation to our existing `birds` NeXML file using the
`add_meta` function.  Because we do not specify a level, it is added to
the root node, referring to the NeXML file as a whole.

```{r}
birds <- add_meta(modified, birds) 
```

The built-in vocabularies are just the tip of the iceberg of established
vocabularies. Here we add an annotation from the `skos` namespace which
describes the history of where the data comes from:

```{r}
history <- meta(property = "skos:historyNote",
  content = "Mapped from the bird.orders data in the ape package using RNeXML")
```

Because `skos` is not in the current namespace list, we add it with a
url when adding this meta element.  We also specify that this annotation
be placed at the level of the `trees` sub-node in the NeXML file.

```{r}
birds <- add_meta(history, 
                birds, 
                level = "trees",
                namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#"))
```


For finer control of the level at which a `meta` element is added,
we will manipulate the `nexml` R object directly using S4 sub-setting,
as shown in the supplement.


Much richer metadata annotation is possible. Later we illustrate how
metadata annotation can be used to extend the base NeXML format to
represent new forms of data while maintaining compatibility with any
NeXML parser. The `RNeXML` package can be easily extended to support
helper functions such as `taxize_nexml` to add additional metadata
without imposing a large burden on the user.


## Reading NeXML metadata

A call to the `nexml` object prints some metadata summarizing the data structure: 

```{r }
birds
```

We can extract all metadata pertaining to the NeXML document as a whole
(annotations of the XML root node, `<nexml>`) with the command

```{r}
meta <- get_metadata(birds) 
```

This returns a named list of available metadata. We can see the kinds
of metadata recorded from the names (showing the first 4):

```{r}
names(meta)[1:4]
```

and can ask for a particular element using the standard list sub-setting
mechanism (i.e. either the name of an element or its numeric position),

```{r}
meta[["dc:title"]]
```

All metadata terms must belong to an explicit *namespace* or vocabulary
that allows a computer to interpret the term precisely. The prefix (before
the `:`) indicates to which vocabulary the term belongs, e.g. `dc` in
this case.  The `get_namespaces` function tells us the definition of
the vocabulary using a link:

```{r}
prefixes <- get_namespaces(birds)
prefixes["dc"]
```


<!-- the citations run off the page -->
Common metadata can be accessed with a few dedicated functions:

```{r get_citation}
get_citation(birds)
```


```{r get_taxa}
get_taxa(birds)
```

Which returns text from the otu element labels, typically used to define
taxonomic names, rather than text from explicit meta elements.



We can also access metadata at a specific level (or use `level=all`
to extract all meta elements in a list).  Here we show only the first
few results:

```{r} 
otu_meta <- get_metadata(birds, level="otu")
otu_meta[1:4]
```

This approach does not preserve the relationship between the metadata element and the 
`otu` it was annotating, and as such is most useful for quick visual exploration. 
Metadata values for specific elements or sets of elements can always be accessed using
the S4 sub-setting mechanisms (see section below).  



## RDF manipulation and SPARQL queries 

The most powerful way to take advantage of the machine-readable nature
of the metadata involves the construction of SPARQL queries. This
approach can exploit the reasoning and logical deduction capacity of
formal vocabularies (ontologies). For a proper introduction to SPARQL
queries, see @W3C_2014 or @Hartig_2012).  

The real expressiveness of SPARQL is better illustrated through more
complex examples such as the one shown in the supplementary materials
(which can also be run from R using the command: `demo("sparql", "RNeXML")`).
This demonstration constructs a cladogram based on the taxonomic ranks
provided in the NeXML. SPARQL allows us to make explicit use of ontological
concepts such as manipulating sub-classes (i.e. that "Genus" is a subclass 
of "Family"), or joining multiple predicates (i.e. retrieve the NCBI identifier
of the OTU that has taxon rank of "Order").  


## Extending the NeXML standard through metadata annotation.

No data standard can anticipate all the uses and extensions researchers
may one day need. Researchers have repeatedly adapted the existing NEXUS
file format to accommodate encoding new kinds of information, but in the
process break compatibility between different flavors of NEXUS files, with
few reliable methods to distinguish between these flavors.  NeXML provides
a solution to this problem through the use of metadata annotation.

Here we illustrate this process using the example of stochastic character
mapping [@Huelsenbecki_2003]. A stochastic character map is simply
an annotation of the branches on a phylogeny, assigning each section
of each branch to a particular "state" (typically of a morphological
characteristic).

@Bollback_2006 provides a widely used stand-alone software implementation
of this method in the software `simmap`, which modified the standard
Newick tree format to express this additional information. This can
break compatibility with other software [^3], and creates a format that
cannot be interpreted without additional information describing this
convention.  By contrast, the NeXML extension is not only backwards
compatible but contains a precise and machine-readable description of
what it is encoding.

In this example, we illustrate how the additional information required
to define a stochastic character mapping (a `simmap` mapping) in NeXML.  

@Revell_2012 describes the `phytools` package for R, which includes
utilities for reading, manipulating, and writing `simmap` files in R.
In this example, we also show how to define `RNeXML` functions that 
map the R representation used by Revell (an extension of the `ape` class)
into the NeXML extension we have defined by using `RNeXML` functions.

Since a stochastic character map simply assigns different states to
parts of a branch (or edge) on the phylogenetic tree, we can create
a NeXML representation by annotating the `edge` elements with appropriate
`meta` elements.  These elements need to describe the character state
being assigned and the duration (in terms of branch-length) that the edge 
spends in that state (Stochastic character maps are specific to time-calibrated
or ultrametric trees).  

NeXML already defines the `characters` element to handle discrete character traits (`nex:char`)
and the states they can assume (`nex:state`).  We will thus reuse the `characters` element for
this purpose, referring to both the character trait and the states by the ids assigned to them
in that element.  (NeXML's convention of referring to everything by id permits a single canonical
definition of each term, making it clear where additional annotation belongs).  For each edge, we 
need to indicate:

<!-- Ugh, do we need to explain that everything in NeXML uses an id so that it can be reused in this way?
     -->

- That our annotation contains a stochastic character mapping reconstruction
- Since many reconstructions are possible for a single edge, we give each reconstruction an id
- We indicate for which character trait we are defining the reconstruction 
- We then indicate which states the character assumes on that edge. 
  For each state realized on the edge, that involves stating: 
    + the state assignment
    + the duration (length of time) for which the edge spends in the given state
    + the order in which the state changes happen (Though we could just assume 
      state transitions are listed chronologically, NeXML suggests making all 
      data explicit, rather than relying on the structure of the data file to
      convey information).  

Thus the annotation for an edge that switches from state `s2` to state 
`s1` of character `cr1` would be constructed like this:

```{r}
 m <- meta("simmap:reconstructions", children = c(
        meta("simmap:reconstruction", children = c(

          meta("simmap:char", "cr1"),
          meta("simmap:stateChange", children = c(
            meta("simmap:order", 1),
            meta("simmap:length", "0.2030"),
            meta("simmap:state", "s2"))),
          
          meta("simmap:char", "cr1"),
          meta("simmap:stateChange", children = c(
            meta("simmap:order", 2),
            meta("simmap:length", "0.0022"),
            meta("simmap:state", "s1")))
          ))))
```

Of course writing out such a definition manually becomes tedious quickly. Because
these are just R commands, we can easily define a function that can loop over an
assignment like this for each edge, extracting the appropriate order, length and
state from an existing R object such as that provided in the `phytools` package.  
Likewise, it is straightforward to define a function that reads this data using
the `RNeXML` utilities and converts it back to the `phytools` package. The full
implementation of this mapping can be seen in the `simmap_to_nexml()` and the
`nexml_to_simmap()` functions provided in the `RNeXML` package.  

As the code indicates, the key step is simply to define the data in meta elements. In 
so doing, we have defined a custom namespace, `simmap`, to hold our variables.  This
allows us to provide a URL with more detailed descriptions of what each of these 
elements mean:

```{r}
nex <- add_namespaces(c(simmap = "https://github.com/ropensci/RNeXML/tree/master/inst/simmap.md"))
```

At that URL we have posted a simple description of each term. 

Using this convention we can generate NeXML files containing `simmap`
data, read those files into R, and convert them back into the `phytools`
package format. These simple functions serve as further illustration of
how `RNeXML` can be used to extend the NeXML standard.  We illustrate
their use briefly here, starting with  loading a `nexml` object containing
a `simmap` reconstruction into R:


```{r}
data(simmap_ex)
```

The `get_trees` function can be used to return an `ape::phylo` tree as
usual.  `RNeXML` automatically detects the `simmap` reconstruction data
and returns includes this in a `maps` element of the `ape::phylo` object,
for use with other `phytools` functions.

```{r}
phy <- nexml_to_simmap(simmap_ex)
```

We can then use various functions from `phytools` designed for `simmap`
objects [@Revell_2012], such as the plotting function:

```{r Figure1, fig.cap="Stochastic character mapping on a phylogeny, as generated by the phytools package after parsing the simmap-extended NeXML."}
library("phytools")
plotSimmap(phy)
```

Likewise, we can convert the object back in the NeXML format and write
it out to file to be read by other users. 

```{r}
nex <- simmap_to_nexml(phy) 
nexml_write(nex, "simmap.xml")
```

Though other NeXML parsers (for instance, for Perl or Python) have
not been written explicitly to express `simmap` data, those parsers will
nonetheless be able to successfully parse this file and expose the `simmap`
data to the user.





## Publishing NeXML files from R

Data archiving is increasingly required by journals in evolutionary biology and 
biodiversity (e.g. @Whitlock_2010).  The burden of preparing and submitting properly
annotated data to archives continues to be a significant barrier [@Tenopir_2011; @Stodden_2014],
and many phylogenetic trees upon which studies are based are inaccessible or lost 
to the research community [@Drew_2013].  `RNeXML` seeks to lower these barriers
by providing functions to immediately archive NeXML files in scientific repositories.  




```{r eval = FALSE}
doi <- nexml_publish(birds, visibility = "public", repository="figshare")
```

Robust data archiving need not involve immediate release of the data. The `figshare` 
repository supports the option of `visibility = "private"`, allowing secure 
publication that provides a way to backup and share data with collaborators 
prior to public release.  


# Software reuse and support

The `RNeXML` package is available through the CRAN repository under
a CC0 license. At the time of writing, the current version is `r packageDescription("RNeXML", fields="Version")`. 
The current development
branch can be found on Github at [https://github.com/ropensci/RNeXML].
See the NEWS file for changes in recent versions. Users are encouraged
to submit bug reports or feature requests in the issues log.  Contact
`ropensci-discuss@googlegroups.com` or the package maintainer for help,
or email the phylogenetics R users group list at `r-sig-phylo@r-project.org`.  


There are many promising directions for further development of the
`RNeXML` package.  The project's [Milestones page](https://github.com/ropensci/RNeXML/issues/milestones) page provides a list of
steps currently planned or under consideration.  In particular, these
milestones include increased support for interactive mode of entering data
(wizards), conversions to and from other existing R phylogenetics formats
such those found in the `phylobase` package [@phylobase],
and support for integration of semantics defined through the
Web Ontology language (OWL), as discussed in the Challenges section.



## Acknowledgements
<!-- FIXME Rutger, Hilmar, add whatever you'd like to acknowledge here -->. 

This project was supported in part by a grant from the Alfred P Sloan
Foundation Grant 2013-6-22 (CB and SC), NSF Grant DBI-1306697 (CB), ... 
We thank F. Michonneau for helpful comments on an earlier version
of this manuscript.
`RNeXML` is part of the rOpenSci project, [http://ropensci.org].



```{r cleanup, include=FALSE, cache=FALSE}
unlink("simmap.xml")
unlink("birds.xml")
unlink("geospiza.xml")
unlink("rdf_meta.xml")
```


<!-- Footnotes -->

[^1]: By providing the optional `dependencies` argument, R will install
additional packages used in some of the advanced examples shown later
but which are not required for basic functionality.

[^2]: For convenience `nexml_read` is also aliased as `read.nexml`,
for users familiar with the somewhat analogous `read.nexus` function.

[^3]: By using the commenting mechanism of the Newick format, it is
possible that other software that doesn't also use the comment mechanism
for some other such purpose would be able to successfully parse the tree.
However there is no way to guarantee that this is the case or for the
data format to describe its use.







